\abstract

In this work we explore the available compositional methods for design of digital circuits.
First we consider the existing method of handshake circuit optimisation via control path 
resynthesis using Petri nets, an approach using structural composition. 
In that approach labelled Petri net parallel composition plays an important role
and we introduce an improvement to the parallel composition algorithm, reducing the number of redundant places in 
the resulting Petri net representations. The proposed algorithm applies to labelled Petri nets in general
and can be applied outside of the handshake circuit optimisation use case.

Next we look at the conditional partial order graph (CPOG) formalism, an approach that allows for a convenient representation of systems
consisting of multiple alternative system behaviours, a phenomenon we call behavioural composition.
We generalise the notion of CPOG and identify an algebraic structure on a more general notion of parameterised graph.
This allows us to do equivalence-preserving manipulation of graphs in symbolic form, which simplifies
specification and reasoning about systems defined in this way, as displayed by two case studies.

As a third contribution we build upon the previous work of CPOG synthesis used 
to generate binary encoding of microcontroller instruction sets and design the corresponding instruction decoder logic. 
The proposed CPOG synthesis technique solves the optimisation problem for the
general case, reducing it to boolean satisfiability problem and uses existing 
SAT solving tools to obtain the result.

\input{Introduction}
\input{Background}
\input{Approach}

\input{par_comp/par_comp}

\input{agda-processed-body}

%% \section{Redundancy in the Petri Nets produced by composition}

%% \section{Eliminating redundancy}

%% \section{Experimental results}

%% \section{Conclusions}

%% \chapter{CPOG Encoding}

\input{pg-encoding/2010_encoding_iet}

% Encoding

% \section{The problem of choosing the optimal encoding}

% \section{Formulating the encoding problem as SAT problem}

% \section{Experimental results}

% \section{Conclusions}

% \chapter{Parametrised Graphs Algebra}

% Algebra

% \section{Generalising CPOGs}

% \section{Definitions}

% \section{Properties and their proofs}

% \section{Formally verified tools for PG formula manipulation}

% \section{Conclusions}

\input{Conclusions}

\chapter*{Appendix}

What follows is the full Agda source code of formal proofs used in Chapter~\ref{chap:PGAlgebra}.

\input{agda-listing-body}

% \section{Workcraft framework plugin: Breeze handshake circuits}

% \section{Workcraft framework plugin: CPOG}

\section{Detailed proofs of PG Algebra properties}

